# Dplyr

```{r 04_setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(eval = T, message = F, warning = F)
```

```{r message=F}
library(tidyverse)
```


## Data Frame

```{r 04_dataframe}
tb_df <- data.frame(
  id = 1:5,
  height = c(1.7, 1.7, 1.8, 1.9, 2.0),
  weight = c(70 , 73 , 80 , 100, 95)
)
tb_df

# Derived variable: Body Mass Index
tb_df$bmi <- tb_df$weight / (tb_df$height)^2
tb_df
```



## Tibble

```{r 04_tibble}
## NSE: Non Standard Evaluation
tb <- tibble(
  id = 1:5,
  height = c(1.7, 1.7, 1.8, 1.9, 2.0),
  weight = c(70 , 73 , 80 , 100, 95),
  # Derived variable: Body Mass Index
  bmi    = weight / height^2
)

tb

```

### Data Frame to Tibble conversion

```{r 04_df_to_tibble}
## cast to tbl
iris_tbl <- as_tibble(iris)
iris_tbl
```

Save column names:

```{r}
## rownames as a column
rownames_to_column(mtcars, var = 'car')

mtcars_tbl <- as_tibble(rownames_to_column(mtcars, var = 'car'))
mtcars_tbl

```



## Data import


### Read a CSV file


CSV: comma separated values file. The first row contains the column names:

```{bash}
head data/dc-wikia-data.csv
```


```{r}
# with a specific delimiter
dc <- read_delim("data/dc-wikia-data.csv", delim = ",")

# with a standard delimiter
dc <- read_csv("data/dc-wikia-data.csv")

# get the specifications used to read the csv
spec(dc)

# with a predefined column type
dc <- read_csv('data/dc-wikia-data.csv',
               col_types = cols(
                 page_id = col_integer(),
                 name = col_character(),
                 urlslug = col_character(),
                 ID = col_factor(),
                 ALIGN = col_factor(),
                 EYE = col_factor(),
                 HAIR = col_factor(),
                 SEX = col_factor(),
                 GSM = col_character(),
                 ALIVE = col_character(),
                 APPEARANCES = col_double(),
                 `FIRST APPEARANCE` = col_character(),
                 YEAR = col_double()
               ))

```

### Column names

Column names standardization: all lower case.

```{r 04_column_names}
# Extraction
old_colnames <- colnames(dc)
old_colnames

# manipulation
new_colnames <- tolower(old_colnames)
new_colnames

# substitution
colnames(dc) <- new_colnames

# result
colnames(dc) 
```

or a short version for the whole process:

```{r}
colnames(dc) <- tolower(colnames(dc))
```


### Save to Rds format

```{r eval=FALSE}
# write to Rds
saveRDS(dc, "data/dc-wikia-data.Rds")
# read from Rds
dc <- readRDS("data/dc-wikia-data.Rds")
```


## Base operations

### Simple selection

```{r}
# select columns: name, appearances, sex
dc_2 <- select(dc, name, appearances, sex)

dc_2
```


### Simple filter

```{r}
# get the "Female Characters"
dc_2_female <- filter(dc_2, sex == "Female Characters")

```

### Pipe 

Two operations in one:

```{r}
filter(select(dc, name, appearances, sex), 
       sex == "Female Characters")
```

Two operations in one pipeline:

```{r}
dc %>%
  select(name, appearances, sex) %>% 
  filter(sex == "Female Characters")
```


RStudio shortcut to write the *pipe* ` %>% `: CTRL + SHIFT + m


### Column Selection (select)

#### Basic selections

```{r}
# select a range
dc %>% 
  select(name, id:year) 

# unselect
dc %>% 
  select(-page_id, -urlslug)
```


```{r}
# rename column 'id' to 'secret_id'
dc %>% 
  rename(secret_id = id)
```

#### Name matching

```{r}
colnames(dc)
```

```{r}
# match the name
dc %>% 
  select(name, contains("appearance"))

dc %>% 
  select(ends_with("id"))

dc %>% 
  select(starts_with("id"))

dc %>% select(matches("appearance.*"))
```

#### Select by type (where)

```{r}
dc %>% select(where(is.factor))

dc %>% select(where(is.character))
```



### Unique values (distinct)

```{r}
dc %>% 
  distinct(id)

dc %>% 
  select(id, sex) %>% 
  distinct()
```


`distinct` is similar to the base R function `unique`, but the latter works on vectors, not on data frames.


### Row Filter (filter)

```{r}
# Single condition
# Example: 
# get "Female Characters" only
dc %>% filter(sex == "Female Characters")

# Match multiple conditions (all of them)
# Example: 
# get the "Female Characters" AND with 'Blond Hair'
dc %>% filter(sex == "Female Characters", hair == "Blond Hair")
# or:
dc %>% filter(sex == "Female Characters" & hair == "Blond Hair")

# Match one of the alternatives (any of them)
# Example: 
# get the "Female Characters" OR with 'Blond Hair'
dc %>% filter(sex == "Female Characters"  | hair == "Blond Hair")
```


```{r}
# Conditions on numbers
# Example: 
# get characters that appeared a greater or equal number of times than 1000 
dc %>% filter(appearances >= 1000)

# Example: 
# within a given interval (inclusive)
dc %>% filter(900 <= appearances, appearances <= 1000)
# or
dc %>% filter(between(appearances, 900, 1000))

# Example: 
# outside a given interval
dc %>% filter(appearances < 900 | 1000 < appearances)
# or
dc %>% filter( ! between(appearances, 900, 1000))

```


```{r}
# set-in operator

# Example: 
# 'appearances' has a value in a vector of possible real numbers
dc %>% filter(appearances %in% c(900, 1000))

# Example: 
# 'hair' has a value in a vector of possible strings
dc %>% filter(hair %in% c('Violet Hair', 'Pink Hair'))
```


```{r}
# * FILTER EXTRAS 

# take rows by their position index
dc %>% slice(5:10)

# Extract a sample of 'n' lines
dc %>% sample_n(3)

# Extract a fraction of all lines
dc %>% sample_frac(0.1)

# Re-sample repeating lines
dc %>% sample_frac(1.5, replace = TRUE)

# if you want to fix the random state of the random sampling
set.seed(123)
dc %>% sample_n(3)
```


### Row sorting (arrange)

Sort using a column as a criteria:

```{r}
dc %>%
  select(name, appearances, sex) %>% 
  filter(sex == "Female Characters") %>% 
  arrange(appearances)
```



```{r}
# sort by one column
dc_2 %>% arrange(appearances)

# sort by one column in descending order
dc_2 %>% arrange(desc(appearances))

# sort by two criteria (the first has the precedence)
dc %>% arrange(sex, desc(appearances))
```


## Transform columns (mutate)

### Create a new column

```{r}
dc %>% 
  mutate(age = 2021 - year) %>% 
  arrange(desc(age)) %>% 
  select(name, age)
```


```{r}
# install.packages("nycflights13")
library(nycflights13)

flights %>% 
  select(dep_delay, arr_delay) %>% 
  mutate(time_gain = dep_delay - arr_delay)
```


### Segment data values into bins

In order to create a categorical variable from a continuous variable you need to segment and sort data values into bins.
You can do that with the `case_when` or the `cut` functions.


#### 'cut' Example


Here a `cut` example. `cut` splits a variable in classes, each of them correspond to an interval of values.

You can automatically split the range of `year` in a number of classes:

```{r}
# You can automatically split the range of `year` in a number of classes:
dc_2 <- dc %>% 
  mutate(classe = cut(year, breaks = 10)) %>% 
  select(name, year, classe)
dc_2
```

Or you can provide a set of separators (`breaks`) and labels for the intervals. Mind the number of breaks should be one more than the labels.

```{r}
# separa le classi sulla variabile separatori e attribuisci la corretta etichetta.
breaks <- 0:10 * 10L + 1900L
labels <- paste('years', as.character(0:9 * 10L))
dc_2 <- dc %>% 
  mutate(classe = cut(year,
                      breaks = breaks, 
                      labels = labels)) %>%
  select(name, year, classe)
dc_2
```


Detail. Here an extract of the procedure, the cut function applied on a single number:

```{r}
cut(1939,
    breaks = breaks, 
    labels = labels)
# 1939 in in the interval between the 4th and 5th breaks
between(1939, breaks[4], breaks[5])
# the selected labels will be the 4th
labels[4]
```


Documentation:

```{r}
help(cut)
```



#### `case_when` example

```{r}
flights %>% 
  select(arr_delay) %>% 
  mutate(delay_class = case_when(
    arr_delay > 1000 ~ "big-delay",
    arr_delay < 1000 & arr_delay > 0 ~ "delay",
    arr_delay <= 0 ~ "no-delay",
    TRUE ~ NA_character_
  ))
```





## Aggregate rows (summarise)

### Scalar-returning aggregations

`summarise` works with functions that return a scalar:

$$
R^n \to  R
$$

For example, the `mean` function take a vector and returns a single value

```{r}
mean(flights$arr_delay, na.rm = TRUE)
```


### Aggregate to a scalar

Prepare the dataset:

```{r}
# install.packages('nycflights13')
library(nycflights13)

flights_tiny <- 
  flights %>% 
  select(dep_delay, arr_delay, carrier, origin, dest)
flights_tiny
```


Calculate a single aggregation:

```{r}
flights_tiny %>% 
  summarise(mean_arr_delay = mean(arr_delay, na.rm = TRUE))
```

Calculate multiple aggregations:

```{r}
flights_tiny %>% 
  summarise(mean_arr_delay   = mean(arr_delay, na.rm = TRUE),
            sd_arr_delay     = sd(arr_delay, na.rm = TRUE),
            mean_dep_delay   = mean(dep_delay, na.rm = TRUE),
            median_arr_delay = median(arr_delay, na.rm = TRUE),
            first_quartile_arr_delay  = quantile(arr_delay, probs = 0.25, na.rm = TRUE),
            second_quartile_arr_delay = quantile(arr_delay, probs = 0.5, na.rm = TRUE))
```



### Vector-returning aggregations

An example is the `quantile` function that return a vector of length equals to `probs`:

```{r}
vec <- 1:11
vec
quantile(vec, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
```

Here the application to `arr_delay`:

```{r}
arr_delay <- flights_tiny$arr_delay

# R^n -> R^m
quantile(arr_delay, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
```

### Aggregate to a vector

`summarise` works also with functions that return a vector of a length different from the starting length. (If the function returns a vector of the same length you will probably want to use `mutate`)

```{r}
# summarise restitutisce questa volta un tibble di 3 righe
flights_tiny %>% 
  summarise(quartile_value = quantile(arr_delay, probs = c(0.25, 0.5, 0.75), na.rm = TRUE), 
            quartile_key   = names(quartile_value)) %>% 
  select(quartile_key, quartile_value)
```



## Compute per groups

### Aggregation per groups

Aggregate per a single variable group:

```{r}
flights %>% 
  group_by(carrier) %>%
  summarise(mean_arr_delay = mean(arr_delay, na.rm = TRUE)) 
```

That aggregation has been made once per each of these groups:

```{r}
flights %>% 
  distinct(carrier)
```



Aggregate per groups generated by multiple variables:

```{r}
flights %>% 
  group_by(carrier, origin, dest) %>%
  summarise(mean_arr_delay = mean(arr_delay, na.rm = TRUE)) %>% 
  ungroup()
```


That aggregation has been made once per each of these groups:

```{r}
flights %>% 
  distinct(carrier, origin, dest)
```



### Grouping attribute

Group is an attribute into the data frame, not a real split. It can be removed with the `ungroup()` function.

```{r}
# summarise remove only one grouping variable by default
# you can read the 'groups' attribute in the tibble
tbl <- flights %>% 
  group_by(carrier, origin, dest) %>%
  summarise(mean_arr_delay = mean(arr_delay, na.rm = TRUE)) 

# before doing another aggregation you may want to 'ungroup()'
tbl %>% 
  ungroup() %>% 
  summarise(mean_arr_delay = mean(mean_arr_delay, na.rm = T))
```


```{r}
# here you see that the grouping attribute can be stored with the tibble variable itself
grouped_flights <- flights %>% 
  group_by(carrier) 

grouped_flights %>% 
  summarise(mean_arr_delay = mean(arr_delay, na.rm = TRUE))
```



### Transformation per group

```{r}
df <- tibble(
  warehouse = rep(LETTERS[1:5], times = 5),
  time = rep(1:5, each = 5),
  total_quantity = as.integer(runif(5 * 5, min = 1, max = 100))
) %>% 
  arrange(warehouse, time)
df
```

- `warehouse` id of the facility where quantity is stored.
- `time` the number of the day for example
- `total_quantity` the number of items stored in the warehouse at that time.

The task now, is to find the `variation` of the number of items per day:

```{r}
df %>% 
  group_by(warehouse) %>% 
  mutate(
    previous_total_quantity = lag(total_quantity),
    variation = total_quantity - previous_total_quantity)
```

Remarks:

- every day 1 the previous_total_quantity is `NA`, and therefore the `variation` (because it is the difference of a number and a `NA`)
- `group_by` is mandatory to keep different `warehouse`s separated each others.





## Combine multiple data frames


### Joins

Joins are a way to merge table based on the correspondence on a key.

```{r}
library(tidyverse)
```

Let us a couple of example tables. Let us define a [foreign-key](https://en.wikipedia.org/wiki/Foreign_key) a column whose values have a correspondence in another table. This creates a relationship among two tables.

```{r}
# Main table, where
# 'id' id number
# 'lower' a foreign key
main_tbl <- tibble(
  id = c(1:2, 3, 4:6), #1:5,
  lower = c(letters[1:2], 'non-letter', letters[4:6])
)
main_tbl
```


And another that contains (usually all) the occurrences of the foreign-key:

```{r}
# The table of letters 
# 'lower' list all the lower case letters
# 'upper' an attribute to the letters, for the example case a upper case copy of the letter
letter_tbl <- tibble(
  lower = letters,
  upper = LETTERS
)
letter_tbl
```


```{r}
# full-join

main_tbl %>% 
  full_join(letter_tbl)
```


```{r}
# inner-join

main_tbl %>% 
  inner_join(letter_tbl)
```


```{r}
# left-join

main_tbl %>% 
  left_join(letter_tbl)
```


```{r}
# right-join

main_tbl %>% 
  right_join(letter_tbl)

# inverting the order a `right_join` return the same as a `left_join`
letter_tbl %>% 
  right_join(main_tbl)


```



### Split and Unite data frames

#### Bind rows

```{r}
mtcar_list <- split(mtcars, as.character(mtcars$cyl))
mtcars_cyl_6 <- mtcar_list[["6"]]
mtcars_cyl_8 <- mtcar_list[["8"]]
```



These two data frames have the same schema, and their rows can be concatenated:

```{r}
bind_rows(mtcars_cyl_6, mtcars_cyl_8)
```



#### Bind columns

```{r}
mtcars_mpg_cyl <- select(mtcars, mpg, cyl)
mtcars_disp_hp_drat <- select(mtcars, hp, drat, wt)
```

These two dataframes have the same number of rows, therefore their columns can be binded:

```{r}
bind_cols(mtcars_disp_hp_drat, mtcars_mpg_cyl)
```

